### 2023-04-09 main/chapter_experiment.tex
add rf cavities

% does not stand alone
The LHC does not stand alone.
It implements the final boosts and collisions of particles, but requires
earlier stages of CERN's systems to inject the protons at suitably high
energies for it to continue their acceleration.
In brief, the accelerator chain flows as follows:
% Accelerator complex
\begin{enumerate}
% where the protons come from
\item The protons come from hydrogen gas, which is initially ionised into a
plasma of protons and $e^-$ ions in the strong electric fields of a
duoplasmotron ion source, in which the protons are initially accelerated by
electric fields towards a cathode~\cite{lhc50years2013}.
% linac2
\item After ionisation, a linear accelerator named Linac2 provides the first
major proton acceleration in the chain, reaching $50\,\eV[M]$
energies~\cite{lhc50years2013}.
% PSB
\item Secondly, the Proton Synchrotron Booster (PSB) accelerates these protons
to $1.4\,\eV[G]$ energies.
The PSB comprises four rings with $160\,\mathrm{m}$
circumference~\cite{linac4protonsource,lhc50years2013}.
% PS
\item Thirdly, protons from the PSB are accelerated by the
Proton Synchrotron (PS) to $25\,\eV[G]$ energies~\cite{lhc2000psInjector}.
TODO
% SPS
\item SPS
% LHC
\item LHC
\end{enumerate}
This sequence is a subset of CERN's full accelerator complex,
which is illustrated in
Figure~\ref{fig:experiment_cern_complex}~\cite{Haffner:1621894}.
Each of these accelerators was upgraded from its previous operations to achieve
the performance demanded by designs for the LHC's high energy and luminosity.

Note that this sequence describes the LHC accelerator complex as it was for
operations that generated the data used in this thesis, in `Run~2' of the LHC.
It has since been upgraded.
For example, since the year 2020 (for Run~3 onwards), Linac2 is replaced by a
new linear accelerator named Linac4,
which accelerates $\mathrm{H}^-$ ions to $160\,\eV[M]$;
these $\mathrm{H}^-$ ions are later stripped of their electrons,
to bare protons by passage through a carbon foil.
Linac4 will supply higher luminosities for future operations of the accelerator
complex~\cite{linac4techreport,linac4protonsource}.
The PSB is since also upgraded to reach $2\,\eV[G]$ energies.

@@

% the accelerator complex
% - sequence of contributing accelerators and their beam energies
% beampipe restricted by size of lep tunnel
% - required particle count makes antiprotons hard
% - proton-proton same charge requires separate beampipes
% - size means most cost efficient to put both beams in one cryostat
% - beampipes therefore mechanically and magnetically coupled
% - superconducting magnets at around 2 to 4 Kelvin
% - cooled with superfluid helium
% - low heat capacity, danger of heating from beam radiation
% Radiofrequency cavities
% - radio waves at 400 MHz such that slow protons accelerated, fast decelerated
% - in LHC, accelerates from 450 GeV to 6.5 TeV (7 target)
% magnets
% - dipole for steering
% - quadrupole for focussing (pairs)
% - higher orders why? septupole at beam entry
% TODO size of beam
% TODO luminosity and its uncertainty
% TODO pile-up
\TODO{Distribution in the mean number of primary vertices per crossing, where the
variability in that mean arises from differing beam conditions over run-time.}
Figure~\ref{fig:experiment_run_2_mu}

>>>>

% does not stand alone
The LHC does not stand alone.
It implements the final boosts and collisions of particles, but requires
earlier stages of CERN's systems to inject the protons at suitably high
energies for it to continue their acceleration.
In brief, the accelerator chain flows as follows:
% Accelerator complex
\begin{enumerate}
% where the protons come from
\item[0.] The protons come from hydrogen gas, which is initially ionised into a
plasma of protons and $e^-$ ions in the strong electric fields of a
`duoplasmotron' ion source, in which the protons are initially accelerated by
electric fields towards a cathode~\cite{lhc50years2013}.
% linac2
\item After ionisation, a linear accelerator named Linac2 provides the first
major proton acceleration in the chain, reaching $50\,\eV[M]$
energies~\cite{lhc50years2013}.
% PSB
\item Secondly, the Proton Synchrotron Booster (PSB) accelerates these protons
to $1.4\,\eV[G]$ energies.
The PSB comprises four rings with $160\,\mathrm{m}$
circumference~\cite{linac4protonsource,lhc50years2013}.
% PS
\item Thirdly, protons from the PSB are accelerated by the
$628\,\mathrm{m}$-circumference
Proton Synchrotron (PS) to $25\,\eV[G]$ energies~\cite{lhc2000psInjector}.
% SPS
\item Fourthly, protons from the PS are further boosted by the
$6.9\,\mathrm{km}$-circumference
Super Proton Synchrotron (SPS) to $450\,\eV[G]$
energies~\cite{synchroton:445034}.
% LHC
\item Finally, protons from the SPS are injected the LHC, which finishes
their acceleration from $250\,\eV[G]$ to $6.5\,\eV[T]$, achieving the
world-leading $\sqrt{s} = 13\,\eV[T]$ proton-proton collisions that are studied
in this thesis.
\end{enumerate}
This sequence is a subset of CERN's full accelerator complex,
illustrated in Figure~\ref{fig:experiment_cern_complex}~\cite{Haffner:1621894},
which also supports various other engineering and research experiments.
Each of these accelerators (Linac2, PSB, PS, and SPS) was upgraded from its
previous operations to achieve the performance demanded by designs for the
LHC's high energy and luminosity.

Note that this sequence describes the LHC accelerator complex as it was for
proton-proton operations that generated the data used in this thesis,
in `Run~2' of the LHC.
It has since been upgraded.
For example, since the year 2020 (for Run~3 onwards), Linac2 is replaced by a
new linear accelerator named Linac4,
which accelerates $\mathrm{H}^-$ ions to $160\,\eV[M]$;
these $\mathrm{H}^-$ ions are later stripped of their electrons,
to bare protons by passage through a carbon foil.
Linac4 will supply higher luminosities for future operations of the accelerator
complex~\cite{linac4techreport,linac4protonsource}.

@@

The LHC's physical design is strongly influenced by the layout of its tunnel,
which was previously occupied by the Large Electron-Positron Collider (LEP).
Colliders of oppositely-charged particle pairs, such as electrons and positrons
or protons and antiprotons, can accelerate both partners in the same beam-pipe.
The same is not true of like-charged pairs, such as protons against protons.
This means that the LHC must squeeze two beam-pipes into the narrow tunnel
where LEP's one beam-pipe previously sat.
The LHC's solution is to put both beams next to each other, in
the same cryostatic container.
This ``two-in-one'' solution fits both beams into the physical space, but adds
engineering challenges because the two beam channels are mechanically and
magnetically coupled due to their
proximity~\cite{lhcDesignreport2004,lhc2008machine}.

Radiofrequency (RF) cavities are the technology used to accelerate protons in
the LHC.
They operate by setting up electron beams to oscillate at a frequency that
resonates with a specially designed cavity, this producing strong oscillating
electromagnetic fields.
These electromagnetic fields oscillate at $400\,\mathrm{MHz}$ in the LHC, and
must be carefully tuned jointly with the incident proton beams to align in
their timing.
When perfectly timed, the RF cavity will accelerate protons that are less
energetic that the target beam energy, and decelerate any that are more
energetic.
This achieves a dynamically stable acceleration.
The LHC uses $16$ superconducting RF cavities contained in four separate
cryostatic modules~\cite{lhcDesignreport2004}.

% TODO magnets
% - dipole for steering
% - quadrupole for focussing (pairs)
% - higher orders why? septupole at beam entry
% TODO size of beam
% TODO luminosity and its uncertainty
% TODO pile-up
\TODO{Distribution in the mean number of primary vertices per crossing, where the
variability in that mean arises from differing beam conditions over run-time.}
Figure~\ref{fig:experiment_run_2_mu}
